---
title: "Schizophrenia cohort demographics"
author: "Rhodes"
output: html_document
---
Date:  `r date()`  

# Introduction
In this analysis we will restrict the DLPFC dataset to the control and schizophrenic subjects only and examine the demographics of that restricted cohort.

# Load data and libraries
```{r Load_Data, cache=FALSE}
RootDirectory = "~/Manuscript"
setwd(RootDirectory)

load(file="0._Rdata_files/3. DLPFCdataset.Rdata") #This file was written by analysis 3. Contains:
#   DLPFCdata  # A matrix containing the log2-transformed data,
#   Phenotypes # A dataframe containing the subject phenotypes 

IlluminaProbes = read.csv(file="resources/Illumina probes.csv", header = TRUE, row.names = 1) # This is the array documentation from Illumina
```

The cohort consists of `r nrow(Phenotypes)`

# Restrict the dataset to the controls and schizophrenics
```{r Sz_only, cache=FALSE}
# The diagnosis information is in the character vector "Phenotypes$NewDx"
Dx = Phenotypes$NewDx
Phenotypes = Phenotypes[(Dx == "Control") | (Dx == "Schizo"),]
Data = DLPFCdata[,(Dx == "Control") | (Dx == "Schizo")]
Dx = as.factor(Dx[(Dx == "Control") | (Dx == "Schizo")])
```

This reduces the size of the cohort to `r nrow(Phenotypes)` individuals.

# Examine the age distribution and restrict the cohort to adults
```{r Age, cache=FALSE}
Age = as.numeric(Phenotypes$AgeDeath)
hist(Age[Dx == "Control"], breaks = 50, col = "red",
     main="Control: red \n Schizophrenic: blue", xlab = "Age")
hist(Age[Dx == "Schizo"], breaks=50, add=T, col=rgb(0, 0, 1, 0.5) )
abline(v = 25, col="black")

# There are 43 subjects for whom "Phenotypes$AgeDeath" is given as "". As a result "Age" is NA for them. We will censor those subjects and the subjects younger than 25.
Discard = Age < 25 | is.na(Age)
Phenotypes = Phenotypes[!Discard,]
Data = Data[,!Discard]
Dx = Dx[!Discard]
Age = Age[!Discard]
```

After restricting the cohorts to adults, the age distribution in the controls and schizophrenics is more or less the same, with a total of `r sum(Dx == "Schizo")` schizophrenics and `r sum(Dx == "Control")` controls.

```{r Age-2, cache=FALSE}
hist(Age[Dx == "Schizo"], 
     breaks = seq(from=20, to=100, by=5), 
     col = "red",
     main="Schizophrenic: red \n Control: blue", 
     xlab = "Age")
hist(Age[Dx == "Control"], add=T,
     breaks = seq(from=20, to=100, by=5), 
     col = rgb(0, 0, 1, 0.5))
```

After restriction of the dataset to subjects 25 y/o or older, the ethnic distribution of the subjects and controls is:
```{r Ethnicity, cache=FALSE, echo=FALSE}
table(Phenotypes$Race, Phenotypes$NewDx)
```

Because there are so few Asians and Hispanics in the cohort we will pool them as "Other".

```{r Race_as_Other, cache=FALSE}
Race = Phenotypes$Race
Phenotypes$Race[(Race == "AS") | (Race == "HISP")] = "Other"
Race = as.factor(Phenotypes$Race)
table(Race, Dx)
```

Note that the cohort is not balanced with respect to race. Specifically, the controls are slightly under-represented in the Caucasian population. 

# Examine distribution of the subjects by gender
```{r}
table(Phenotypes$Gender, Dx)
```

As might be expected in a Medical Examiner cohort where the control subjects include accidental death and homocide victims, men are somewhat over-represented in the controls.  

For some reason that imbalance is much more prominent in the Caucasian than African American sub-cohorts.  

### Caucasian only  
```{r}
table(Phenotypes$Gender[Race == "CAUC"], Dx[Race == "CAUC"])
```

### African Americans
```{r}
table(Phenotypes$Gender[Race == "AA"], Dx[Race == "AA"])
```
 
**This gender x race interaction is something which should be kept in mind if either gender or race are critical covariates in the final results.**

# RIN
```{r RIN, cache=FALSE}
Phenotypes$RIN = as.numeric(as.character(Phenotypes$RIN))
hist(Phenotypes$RIN[Dx == "Schizo"], 
     breaks = seq(from=4, to=10, by=0.3), 
     probability = "TRUE",
     col = "red",
     main="Schizophrenic: red \n Control: blue", 
     ylim=c(0,1),
     xlab = "RIN")
hist(Phenotypes$RIN[Dx == "Control"], add=T,
     breaks = seq(from=4, to=10, by=0.3), 
     probability = "TRUE",
     col = rgb(0, 0, 1, 0.5))

t.test(Phenotypes$RIN ~ Phenotypes$NewDx)
t.test(Phenotypes$RIN[Phenotypes$NewDx == "Schizo"], Phenotypes$RIN[Phenotypes$NewDx == "Control"]);
wilcox.test(Phenotypes$RIN[Phenotypes$NewDx == "Schizo"], Phenotypes$RIN[Phenotypes$NewDx == "Control"]);
```

# Format the phenotype dataframe to include only the relavent information and the correct data types.
```{r Format_Pheno, cache=FALSE}
Pheno = Phenotypes[,c("NewDx", "AgeDeath", "Gender", "Race", "CauseDeath", "RIN", "BATCH")]
colnames(Pheno) = c("Dx", "Age", "Gender", "Race", "CauseDeath", "RIN", "BATCH")
Pheno$Dx = as.factor(Pheno$Dx)
Pheno$Age = as.numeric(Pheno$Age)
Pheno$Gender = as.factor(Pheno$Gender)
Pheno$Race = as.factor(Pheno$Race)
Pheno$RIN = as.numeric(Pheno$RIN)
Pheno$BATCH = as.factor(Pheno$BATCH)
```

# Transpose the expression array data
Transpose the expression array data to put it into the format expected by most R functions with the subjects as rows and the probes as columns.
```{r Transpose_data, cache=FALSE}
Data = t(Data)
```

We are now ready to begin the analysis of this limited dataset restricted to the DLPFC data from the controls and schizophrenics.

# Save the results
```{r}
save(file="0._Rdata_files/4. DLPFC-Sz dataset.Rdata", 
     Data,          # The normalized, log2-transformed expression array data restricted to Sz and controls
     Pheno,         # The subject phenotype data 
     IlluminaProbes # The array annotation from Illumina
     )
```


```{r}
sessionInfo()
```
